## **Mapping QC** {#mapqc}

### **BAM mapping QC**

<details>
<summary>Description</summary>

__BAM QC Summary (NGS-bit mapQC)__

This section summarizes sequencing and alignment quality metrics derived
from BAM files. These metrics reflect the overall data quality, mapping
performance, coverage characteristics, and potential technical biases.

A 'WARN' or 'FAIL' interpretation may be considered under the following
general conditions (thresholds can be adjusted based on assay design):


__Read processing & mapping__

* `trimmed base percentage`  
  Fraction of bases removed during read trimming.  
  High values may indicate poor raw read quality or adapter contamination.

* `clipped base percentage`  
  Fraction of bases soft/hard clipped during alignment.  
  Elevated levels may reflect mapping issues or structural variation.

* `mapped read percentage`  
  Percentage of reads successfully aligned to the reference genome.  
  - Expected: >95%  
  - Low values may indicate contamination or poor sequencing quality.

* `properly-paired read percentage`  
  Percentage of read pairs aligned in proper orientation.  
  - Expected: >90%  
  - Low values may indicate library prep issues.

* `duplicate read percentage`  
  Fraction of PCR/optical duplicates.  
  - Expected: <40%  
  - High duplication reduces effective coverage.


__Target enrichment & coverage__

* `on-target read percentage`  
  Fraction of reads mapping to target regions.  
  - Reflects capture efficiency.

* `near-target read percentage`  
  Fraction of reads mapping near target regions.  
  - Useful for evaluating capture specificity.

* `target region read depth`  
  Average sequencing depth across target regions.  
  - Key determinant of variant detection sensitivity.

* `target region read depth (no ol)`  
  Depth excluding overlapping reads.  
  - Provides a more conservative estimate.

* `bases usable (MB)`  
  Total number of usable sequenced bases.  
  - Reflects effective sequencing output.


__Coverage uniformity__

* `target region 10x/20x/30x/50x/60x/100x/200x/500x percentage`  
  Percentage of target bases covered at ≥X depth.  
  - Used to assess coverage completeness.

* `target region half depth percentage`  
  Fraction of regions with depth ≥50% of mean coverage.  
  - Indicates coverage uniformity.


__Fragment & insert size__
* `insert size`  
  Average fragment length.  
  - Deviations may indicate library preparation issues.

__Bias & composition__

* `AT dropout`  
  Underrepresentation of AT-rich regions.  
  - High values suggest sequencing bias.

* `GC dropout`  
  Underrepresentation of GC-rich regions.  
  - High values indicate GC bias affecting coverage.


__Sex chromosome & contamination indicators__

* `chrY/chrX read ratio`  
  Ratio of reads mapping to chromosome Y vs X.  
  - Used for sex inference or contamination detection.

__General interpretation guidance__

* High duplication or low mapping rates → potential library or sequencing issues  
* Low on-target rate → inefficient capture  
* Low coverage or uneven depth → reduced variant calling sensitivity  
* High GC/AT dropout → bias in sequencing chemistry  
* Abnormal chrY/chrX ratio → possible sample swap or contamination  

</details>

```{r bam-qc-tumor-normal, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(tibble)
library(reactable)
library(htmltools)

read_mapqc <- function(file, sample_type) {
  readr::read_lines(file) %>%
    purrr::discard(~ stringr::str_trim(.x) == "") %>%
    tibble::tibble(raw = .) %>%
    tidyr::separate(
      raw,
      into = c("metric", "value"),
      sep = ":\\s*",
      extra = "merge",
      fill = "right"
    ) %>%
    dplyr::mutate(
      metric = stringr::str_trim(metric),
      value = as.numeric(value),
      sample_type = sample_type
    )
}

qc_rules <- function(metric, value) {

  dplyr::case_when(
    metric == "mapped read percentage" & value < 90 ~ "FAIL",
    metric == "mapped read percentage" & value < 95 ~ "WARN",
    
    metric == "properly-paired read percentage" & value < 85 ~ "FAIL",
    metric == "properly-paired read percentage" & value < 90 ~ "WARN",
    
    metric == "on-target read percentage" & value < 50 ~ "FAIL",
    metric == "on-target read percentage" & value < 60 ~ "WARN",
    
    metric == "duplicate read percentage" & value > 60 ~ "FAIL",
    metric == "duplicate read percentage" & value > 40 ~ "WARN",
    
    metric == "target region read depth" & value < 20 ~ "FAIL",
    metric == "target region read depth" & value < 30 ~ "WARN",
    
    metric == "GC dropout" & value > 30 ~ "FAIL",
    metric == "GC dropout" & value > 20 ~ "WARN",
    
    TRUE ~ "PASS"
  )
}

tumor_qc <- read_mapqc(params$ngs_bit_mapqc_tumor, "Tumor")

has_normal <- !is.null(params$ngs_bit_mapqc_normal) &&
  !is.na(params$ngs_bit_mapqc_normal) &&
  file.exists(params$ngs_bit_mapqc_normal)

qc_long <- tumor_qc

if (has_normal) {
  normal_qc <- read_mapqc(params$ngs_bit_mapqc_normal, "Normal")
  qc_long <- dplyr::bind_rows(tumor_qc, normal_qc)
}

# pivot 后保证 Tumor / Normal 列存在性
qc_tbl <- qc_long %>%
  dplyr::mutate(
    is_percent = stringr::str_detect(metric, regex("percentage|dropout", ignore_case = TRUE)),
    group = dplyr::case_when(
      stringr::str_detect(metric, stringr::regex("mapped|paired|duplicate|trimmed|clipped", ignore_case = TRUE)) ~ "Mapping / Reads",
      stringr::str_detect(metric, stringr::regex("target|depth|usable", ignore_case = TRUE)) ~ "Target / Depth",
      stringr::str_detect(metric, stringr::regex("insert", ignore_case = TRUE)) ~ "Fragment",
      stringr::str_detect(metric, stringr::regex("dropout|chrY|chrX|allele", ignore_case = TRUE)) ~ "Bias / Sex / AF",
      TRUE ~ "Other"
    )
  ) %>%
  dplyr::select(group, metric, sample_type, value, is_percent) %>%
  tidyr::pivot_wider(
    names_from = sample_type,
    values_from = value
  ) %>%
  dplyr::arrange(group, metric)

bar_cell <- function(value, is_percent) {
  if (is.null(value) || is.na(value)) return("")
  
  if (isTRUE(is_percent)) {
    width <- max(0, min(value, 100))
    div(
      style = "display:flex;align-items:center;gap:8px;",
      div(
        style = "
          width:110px;
          height:10px;
          background:#e5e7eb;
          border-radius:999px;
          overflow:hidden;
        ",
        div(
          style = paste0(
            "height:100%;",
            "width:", width, "%;",
            "background:#2563eb;",
            "border-radius:999px;"
          )
        )
      ),
      span(sprintf("%.2f%%", value))
    )
  } else {
    span(format(round(value, 2), big.mark = ","))
  }
}

# 只给真实存在的列设置 colDef
qc_columns <- list(
  group = colDef(name = "Category"),
  metric = colDef(
    name = "QC Metric",
    minWidth = 270,
    style = list(fontWeight = 600)
  ),
  is_percent = colDef(show = FALSE)
)

if ("Tumor" %in% colnames(qc_tbl)) {
  qc_columns$Tumor <- colDef(
    name = "Tumor",
    minWidth = 230,
    cell = function(value, index) {
      bar_cell(value, qc_tbl$is_percent[index])
    }
  )
}

if ("Normal" %in% colnames(qc_tbl)) {
  qc_columns$Normal <- colDef(
    name = "Normal",
    minWidth = 230,
    cell = function(value, index) {
      bar_cell(value, qc_tbl$is_percent[index])
    }
  )
}

qc_tbl <- qc_tbl %>%
  mutate(
    Tumor_status = qc_rules(metric, Tumor),
    Normal_status = if ("Normal" %in% colnames(.)) qc_rules(metric, Normal) else NA
  )

status_badge <- function(x) {
  if (is.na(x)) return("")
  
  bg <- case_when(
    x == "PASS" ~ "#16a34a",
    x == "WARN" ~ "#ea580c",
    x == "FAIL" ~ "#dc2626"
  )
  
  span(
    style = paste0(
      "padding:4px 10px;",
      "border-radius:999px;",
      "color:white;",
      "font-weight:700;",
      "font-size:12px;",
      "background:", bg
    ),
    x
  )
}

# div(
#   style = "
#     width: 90%;
#     margin: 20px auto;
#   ",
# reactable(
#   qc_tbl,
#   groupBy = "group",
#     theme = reactableTheme(
#     headerStyle = list(
#       background = "linear-gradient(90deg, #166534, #22c55e)",
#       color = "white"
#     ),borderColor = "#e5e7eb"
#   ),
#   columns = list(
#     metric = colDef(name = "QC Metric"),
#     Tumor = colDef(
#       cell = function(value, index) {
#         tagList(
#           bar_cell(value, qc_tbl$is_percent[index]),
#           status_badge(qc_tbl$Tumor_status[index])
#         )
#       }
#     ),
#     
#     Tumor_status = colDef(show = FALSE),
#     Normal = colDef(
#       show = "Normal" %in% colnames(qc_tbl),
#       cell = function(value, index) {
#         tagList(
#           bar_cell(value, qc_tbl$is_percent[index]),
#           status_badge(qc_tbl$Normal_status[index])
#         )
#       }
#     )
#   )
# ),
# )
```

### **Tumor-Normal checkmate QC**
<details>
<summary>Description</summary>

**Tumor-Normal Concordance QC**

This section evaluates whether the tumor and matched normal samples are likely derived from the same individual.

The concordance value represents the percentage of informative germline markers showing consistent genotypes between the tumor and normal samples. A high concordance supports correct tumor-normal pairing, while a low concordance may suggest sample mismatch, sample swap, contamination, or insufficient sequencing quality at shared markers.

**Interpretation criteria**

- `PASS`: Concordance ≥ 90%; tumor and normal samples are likely matched.
- `WARN`: Concordance < 90%; tumor-normal pairing should be reviewed carefully.

**Additional metrics**

- `Markers used`: Number of informative markers used for concordance calculation.
- `Coverage threshold per marker`: Minimum read depth required for a marker to be included.
- `Minimum mapping quality`: Minimum mapping quality used during marker evaluation.
- `Minimum base quality`: Minimum base quality used during marker evaluation.

</details>
```{r conpair_concordance_qc, echo=FALSE, message=FALSE, warning=FALSE}


has_conpair_concordance <- !is.null(params$conpair_concordance) &&
  !is.na(params$conpair_concordance) &&
  file.exists(params$conpair_concordance)

if (has_conpair_concordance) {

  conpair_lines <- readr::read_lines(params$conpair_concordance)

  concordance_line <- conpair_lines[str_detect(conpair_lines, "^Concordance:")]
  marker_line <- conpair_lines[str_detect(conpair_lines, "^Based on")]
  mapq_line <- conpair_lines[str_detect(conpair_lines, "^Minimum mappinq quality:|^Minimum mapping quality:")]
  baseq_line <- conpair_lines[str_detect(conpair_lines, "^Minimum base quality:")]

  concordance_value <- concordance_line %>%
    str_extract("[0-9.]+(?=%)") %>%
    as.numeric()

  marker_info <- marker_line %>%
    str_extract("[0-9]+/[0-9]+ markers")

  coverage_threshold <- marker_line %>%
    str_extract("(?<=threshold : )[0-9]+ reads")

  min_mapping_quality <- mapq_line %>%
    str_extract("[0-9]+$") %>%
    as.numeric()

  min_base_quality <- baseq_line %>%
    str_extract("[0-9]+$") %>%
    as.numeric()

  concordance_status <- dplyr::case_when(
    is.na(concordance_value) ~ "NA",
    concordance_value >= 90 ~ "PASS",
    concordance_value < 90 ~ "WARN"
  )

  interpretation <- dplyr::case_when(
    is.na(concordance_value) ~ "Concordance value could not be parsed.",
    concordance_value >= 90 ~ "Tumor and normal samples are likely from the same individual.",
    concordance_value < 90 ~ "Tumor and normal samples may not be matched; possible sample swap or contamination should be checked."
  )

  conpair_tbl <- tibble::tibble(
    Metric = c(
      "Tumor-Normal Concordance",
      "Markers used",
      "Coverage threshold per marker",
      "Minimum mapping quality",
      "Minimum base quality",
      "Interpretation"
    ),
    Value = c(
      paste0(sprintf("%.2f", concordance_value), "%"),
      marker_info,
      coverage_threshold,
      as.character(min_mapping_quality),
      as.character(min_base_quality),
      interpretation
    ),
    Status = c(
      concordance_status,
      "",
      "",
      "",
      "",
      concordance_status
    )
  )

  status_badge <- function(status) {
    if (is.na(status) || status == "") return("")

    bg <- dplyr::case_when(
      status == "PASS" ~ "#16a34a",
      status == "WARN" ~ "#f59e0b",
      status == "FAIL" ~ "#dc2626",
      TRUE ~ "#6b7280"
    )

    htmltools::span(
      style = paste0(
        "display:inline-block;",
        "padding:4px 10px;",
        "border-radius:999px;",
        "color:white;",
        "font-weight:700;",
        "font-size:12px;",
        "background:", bg, ";"
      ),
      status
    )
  }

  value_cell <- function(value, index) {
    metric <- conpair_tbl$Metric[index]
    status <- conpair_tbl$Status[index]

    if (metric %in% c("Tumor-Normal Concordance", "Interpretation")) {
      bg <- dplyr::case_when(
        status == "PASS" ~ "#dcfce7",
        status == "WARN" ~ "#fef3c7",
        status == "FAIL" ~ "#fee2e2",
        TRUE ~ "transparent"
      )

      color <- dplyr::case_when(
        status == "PASS" ~ "#166534",
        status == "WARN" ~ "#92400e",
        status == "FAIL" ~ "#991b1b",
        TRUE ~ "#111827"
      )

      htmltools::span(
        style = paste0(
          "display:inline-block;",
          "padding:4px 8px;",
          "border-radius:6px;",
          "font-weight:700;",
          "background:", bg, ";",
          "color:", color, ";"
        ),
        value
      )
    } else {
      htmltools::span(value)
    }
  }

  htmltools::div(
    style = "
      width: 90%;
      margin: 20px auto;
    ",
    reactable::reactable(
      conpair_tbl,
      pagination = FALSE,
      bordered = TRUE,
      striped = TRUE,
      highlight = TRUE,
      compact = TRUE,
      columns = list(
        Metric = colDef(
          name = "QC Metric",
          minWidth = 220,
          style = list(fontWeight = 700)
        ),
        Value = colDef(
          name = "Value",
          minWidth = 350,
          cell = function(value, index) value_cell(value, index)
        ),
        Status = colDef(
          name = "Status",
          minWidth = 100,
          cell = function(value) status_badge(value)
        )
      ),
      theme = reactableTheme(
        borderColor = "#e5e7eb",
        stripedColor = "#f9fafb",
        highlightColor = "#f3f4f6",
        cellPadding = "8px 10px",
        style = list(
          fontFamily = "Arial, Helvetica, sans-serif",
          fontSize = "13px"
        ),
        headerStyle = list(
          background = "#1e3a8a",
          color = "white",
          fontWeight = "700"
        )
      )
    )
  )
}
```

### **Purple CNV calling QC**

<details>
<summary>Description</summary>

__PURPLE Summary__

PURPLE outputs a QC status along with a summary
for the inferred purity and ploidy of the somatic sample.
The 'QC Status' field reflects how we have determined the purity of the sample.
A 'FAIL' or 'WARN' QC status can be attributed to several factors:

* `FAIL_CONTAMINATION`: measured contamination in the tumor (by Amber) is >10%
* `FAIL_NO_TUMOR`: no evidence of tumor is found in the sample, when all
  following criteria are satisfied:
  - tumor has one or more HOTSPOT SV or point mutation
  - SNV sum(allele read count) > 5000
  - SV sum(startTumorVariantFragmentSupport) > 1000 (excluding SGL breakends)
* `WARN_DELETED_GENES`: more than 280 homozygously deleted genes.
  This sometimes occurs in samples with very high MB scale GC bias, and
  particularly affects high GC content regions such as CHR 19.
  It may also indicate a poor fit.
* `WARN_HIGH_COPY_NUMBER_NOISE`: more than 220 copy number segments __not__
  supported at either end by SV breakpoints.
  Indicates samples with extreme GC bias, with differences
  in depth of >= 10x between high and low GC regions. GC normalisation is
  unreliable when the corrections are so extreme so it is recommended to fail
  the sample (concerns with miscalled deletions or amplifications or have
  poor sensitivity in high GC regions).
* `WARN_GENDER_MISMATCH`: if the AMBER and COBALT inferred genders are
  inconsistent then the COBALT one is used but the sample is failed.
* `WARN_LOW_PURITY`: fitted purity < 20%

</details>

```{r funcs}
purple_dir <- params$purple_res_dir # should be relative to the knitting directory
ascat_dir <- params$ascat_res_dir 

blank_lines <- function(n = 10) {
  cat(rep("&nbsp;  ", n), sep = "\n")
}

# get js indices when using in DT options
col_ind_js <- function(dt, col) {
  ntab <- setNames((1:ncol(dt) - 1), names(dt))
  stopifnot(all(col %in% names(dt)))
  sort(unname(ntab[col]))
}
```

```{r purple_prep}
#---- PURPLE QC Table ----#
purple_qc_summary <- dplyr::bind_rows(
  gpgr::purple_qc_read(params$purple_qc)[["summary"]] %>%
    dplyr::filter(!variable %in% c("Purity", "Gender")),
  gpgr::purple_purity_read(params$purple_purity)[["summary"]]
)  %>%  dplyr::mutate(
    value = as.character(value),
    details = as.character(details)
  )
```

```{r purple_qc}
qc_summary_all <- dplyr::tribble(
  ~n, ~variable, ~value, ~details,
  1, "Sample Name", params$tumor_name, "Sample Name of this run",
  # 20, "CNVs (Somatic)", as.character(glue::glue(
  #   "Min: {cnv_summary[['som']]$Min}; ",
  #   "Max: {cnv_summary[['som']]$Max}; ",
  #   "N: {cnv_summary[['som']]$N} "
  # )), "",
  # 17, "SNVs (Somatic)", snv_summary, "Summary SNV counts across main processes, in order",
  # 19, "SVs (melted)", paste(rev(sv_summary$melted), collapse = ", "),
  # "Summary of SVs after splitting annotations for each event into a separate variant row.",
  22, "Genome", params$genome_version, "Human genome assembly used."
) %>%
  dplyr::mutate(
    value = as.character(value),
    details = as.character(details)
  ) %>%
  dplyr::bind_rows(
    purple_qc_summary
  ) %>%
  dplyr::arrange(n) %>%
  dplyr::select(-n)

```
```{r summary_tables}

library(dplyr)
library(reactable)
library(htmltools)

orange <- "#FED8B1"
red <- "#FF4646"
green <- "#dcfce7"
blue_header <- "#1e3a8a"

qc_summary_all2 <- qc_summary_all %>%
  mutate(
    value = as.character(value),
    details = as.character(details)
  )

# 拆成左右两个表
mid <- ceiling(nrow(qc_summary_all2) / 2)

qc_left <- qc_summary_all2 %>%
  slice(1:mid)

qc_right <- qc_summary_all2 %>%
  slice((mid + 1):n())

value_cell <- function(value, index, data) {
  variable <- data$variable[index]
  
  bg <- dplyr::case_when(
    variable == "QC_Status" & grepl("FAIL", value) ~ red,
    variable == "QC_Status" & grepl("WARN_", value) ~ orange,
    variable == "QC_Status" & grepl("PASS", value) ~ green,
    variable == "Hypermutated" & grepl("TRUE", value) ~ orange,
    TRUE ~ "transparent"
  )
  
  span(
    style = paste0(
      "display:inline-block;",
      "padding:3px 8px;",
      "border-radius:6px;",
      "font-weight:700;",
      "background:", bg, ";"
    ),
    value
  )
}

make_qc_table <- function(dat) {
  reactable(
    dat,
    pagination = FALSE,
    bordered = TRUE,
    striped = TRUE,
    highlight = TRUE,
    compact = TRUE,
    columns = list(
      variable = colDef(
        name = "Metric",
        minWidth = 160,
        cell = function(value, index) {
          span(
            title = dat$details[index],
            style = "font-weight:700; cursor:help;",
            value
          )
        }
      ),
      value = colDef(
        name = "Value",
        minWidth = 130,
        cell = function(value, index) value_cell(value, index, dat)
      ),
      details = colDef(show = FALSE)
    ),
    theme = reactableTheme(
      borderColor = "#dbeafe",
      stripedColor = "#eff6ff",
      highlightColor = "#dbeafe",
      cellPadding = "6px 8px",
      style = list(
        fontFamily = "Arial, Helvetica, sans-serif",
        fontSize = "12px"
      ),
      headerStyle = list(
        background = blue_header,
        color = "white",
        fontWeight = "700"
      )
    )
  )
}

# div(
#   style = "
#     width: 90%;
#     margin: 20px auto;
#     display: grid;
#     grid-template-columns: 1fr 1fr;
#     gap: 16px;
#     align-items: start;
#   ",
#   make_qc_table(qc_left),
#   make_qc_table(qc_right)
# )

```

```{r patient_info_table2, echo=FALSE, results='asis'}

```
